Probabilistic inference in dynamical systems

### 2024-04-08 nbs/concepts/inferenceindynamicalsystems/index.qmd
fix(nbs): account for parameter space singularity Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

| Symbol   | Description                 | Units                         | $O(-)$ Estimate | Note                                                                          |
| -------- | --------------------------- | ----------------------------- | --------------- | ----------------------------------------------------------------------------- |
| $u$      | Number of pre-mRNA          | molecules/cell                | $10^0 - 10^4$   | Wide range accounts for low to high gene expression levels.                   |
| $s$      | Number of mRNA              | molecules/cell                | $10^0 - 10^5$   | Similar to $u$; depends on gene expression and stability of mRNA.             |
| $t$      | Time                        | seconds ($s$) to hours ($hr$) | $5s$ - $48hr$   | Depends on the duration of the experimental observation.                      |
| $\alpha$ | Production rate of pre-mRNA | molecules/(cell·hr)           | $10^0 - 10^3$      | Many transcripts are produced at rates in the range or 1 to 1000 per hour.    |
| $\beta$  | Splicing rate of pre-mRNA   | $hr^{-1}$                     | $10^{-1} - 10^2$  | Many Pre-mRNA to mRNA splicing rates are in the range of 1 minute to 6 hours. |
| $\gamma$ | Degradation rate of mRNA    | $hr^{-1}$                     | $10^{-2} - 10^0$     | Many mRNA half-lives are in the range of a half-hour to a day.                |

@@

Given the state variables representing concentrations of pre-mRNA, $u$, 
and mRNA, $s$, we have the following ordinary differential equations

\begin{align}
\frac{du}{dt} & = \alpha - \beta u  \label{eq-dudt} \\
\frac{ds}{dt} & = \beta u - \gamma s  \label{eq-dsdt}
\end{align}

\eqref{eq-dudt} and \eqref{eq-dsdt} are usually presented with 
the concept that the parameter $\alpha$ could better account for the external 
inputs to the regulation of transcription of the gene if it were allowed to be 
a function of time, $\alpha(t)$. We will return to this point later, but 
for now we will assume that $\alpha$ and other parameters are constant.

@@

It is common to perform a dimensional analysis of the model to ensure the
units of the variables and parameters are consistent and reduce the total number
of parameters by the number of dimensions. Intuitively, this folds or maps parameters
to equivalence classes of similar dynamics.  For reference, the procedure
we outline is also commonly referred to as Buckingham's $\Pi$ theorem.

@@

| $\beta$ (hr$^{-1}$) | 1 unit of $t^*$ (hr) | 10 units of $t^*$ (hr) | 100 units of $t^*$ (hr) | 1 unit of $t^{\ast}$ | 10 units of $t^{\ast}$ | 100 units of $t^{\ast}$ |
|---------------------|----------------------|------------------------|-------------------------|----------------------|------------------------|-------------------------|
| $10^{-1}$           | 10                   | 100                    | 1000                    | 1                    | 10                     | 100                     |
| $10^0$              | 1                    | 10                     | 100                     | 1                    | 10                     | 100                     |
| $10^1$              | 0.1                  | 1                      | 10                      | 1                    | 10                     | 100                     |
| $10^2$              | 0.01                 | 0.1                    | 1                       | 1                    | 10                     | 100                     |

@@

and define dimensionelss variables and parameters as

@@

```{python}
#| code-fold: true
du_dt_dimless = Eq(
    beta * diff(u_star * alpha / beta, t_star),
    du_dt.rhs.subs(
        {
            u: u_star * alpha / beta,
            s: s_star * alpha / beta,
            t: t_star / beta,
            gamma: gamma_star * beta,
        }
    ),
).simplify()
du_dt_dimless
```

@@

```{python}
# | code-fold: true
ds_dt_dimless = Eq(
    beta * diff(s_star * alpha / beta, t_star),
    ds_dt.rhs.subs(
        {
            u: u_star * alpha / beta,
            s: s_star * alpha / beta,
            t: t_star / beta,
            gamma: gamma_star * beta,
        }
    ),
).simplify()

@@

\begin{align}
\frac{du^{\ast}}{dt^{\ast}} & = 1 - u^{\ast}  \label{eq-dustardtstar} \\
\frac{ds^{\ast}}{dt^{\ast}} & = u^{\ast} - \gamma^{\ast} s^{\ast}  \label{eq-dsstardtstar}
\end{align}

@@

We'll solve the dimensionless system analytically, focusing on what can be 
understood about the system's dynamics.
Given generic initial conditions $u^{\ast}(0) = u^{\ast}_0$ and $s^{\ast}(0) = s^{\ast}_0$, the 
first equation

@@

$C_1$ is the constant determined by initial conditions, which can be shown to be $C_1 = u^{\ast}_0 - 1$, giving us

@@

This equation describes how the concentration of pre-mRNA, in units of the 
characteristic concentration scale given by $\alpha/\beta$, evolves over the 
timescale characterized by the splicing rate. The term $e^{-t^{\ast}}$ 
indicates that any deviation of $u^{\ast}$ from its steady state $1$ decays 
exponentially over time.

@@

Then the complete solutions for the dimensionless system are given by

@@

@@

The **pre-mRNA dynamics** $u^{\ast}(t^{\ast})$ \eqref{eq-ustar-dimless} quickly 
settles to its steady state value of 1, regardless of the initial concentration, 
reflecting the balance between transcription and splicing rates. Note again that this
does not refer to an arbitrary value of $1$ but rather represents the equivalence
class of ratios $\alpha/\beta$.

The **mRNA dynamics** $s^{\ast}(t^{\ast})$ \eqref{eq-sstar-dimless} reveals a 
slightly more complicated balance of forces. The first transient term shows an exponential 
decay influenced by $\gamma^{\ast}$, the degradation rate relative to the splicing rate. 
The second transient term reflects how the 
initial deviation in pre-mRNA concentration from its steady state affects mRNA 
levels over time. The steady state of $s^{\ast}$ simply reflects the balance
of the splicing rate, which is the only source of mRNA production in this model,
and the degradation rate also given, just as in the case of the pre-mRNA, in units of 
the transcription rate relative to the splicing rate.

@@

```{python}
# | echo: false
# | output: false
%load_ext autoreload
%autoreload 2
```

```{python}
# | label: transcription-splicing-degradation-simulation
from pyrovelocity.models import solve_transcription_splicing_model
from pyrovelocity.models import solve_transcription_splicing_model_analytical
from pyrovelocity.logging import configure_logging
from jax import numpy as jnp

@@

ts0 = jnp.linspace(0.1, 4.0, 40)
ts1 = jnp.linspace(4.0 + (10.0 - 4.0) / 20, 10.0, 20)
ts = jnp.concatenate([ts0, ts1])
initial_state = jnp.array([0.1, 0.1])
params = (0.99,)

solution = solve_transcription_splicing_model(
    ts,
    initial_state,
    params,
)
analytical_solution = solve_transcription_splicing_model_analytical(
    ts,
    initial_state,
    params,
)

@@

plot_deterministic_simulation_trajectories(
    solution=solution,
    title_prefix="TSD Model Simulated",
)
```

We see that if we plot the analytical solution, we get the same result as shown in @fig-transcription-splicing-degradation-plot-analytical-trajectory-1.

```{python}
# | label: fig-transcription-splicing-degradation-plot-analytical-trajectory-1
# | code-fold: true
# | fig-cap: Analytical trajectories of pre-mRNA and mRNA concentrations over time for the transcription-splicing-degradation model.
from pyrovelocity.plots import plot_deterministic_simulation_trajectories

plot_deterministic_simulation_trajectories(
    solution=analytical_solution,
    title_prefix="TSD Model Analytical",
)
```

and the phase portraits as shown in @fig-transcription-splicing-degradation-plot-simulation-phase-portrait-1.

@@

plot_deterministic_simulation_phase_portrait(
    solution=solution,
    title_prefix="TSD Model",
)
```

>>>>

| Symbol   | Description                 | Units                         | $O(-)$ Estimate  | Note                                                                          |
| -------- | --------------------------- | ----------------------------- | ---------------- | ----------------------------------------------------------------------------- |
| $u$      | Number of pre-mRNA          | molecules/cell                | $10^0 - 10^4$    | Wide range accounts for low to high gene expression levels.                   |
| $s$      | Number of mRNA              | molecules/cell                | $10^0 - 10^5$    | Similar to $u$; depends on gene expression and stability of mRNA.             |
| $t$      | Time                        | seconds ($s$) to hours ($hr$) | $5s$ - $48hr$    | Depends on the duration of the experimental observation.                      |
| $\alpha$ | Production rate of pre-mRNA | molecules/(cell·hr)           | $10^0 - 10^3$    | Many transcripts are produced at rates in the range or 1 to 1000 per hour.    |
| $\beta$  | Splicing rate of pre-mRNA   | $hr^{-1}$                     | $10^{-1} - 10^2$ | Many Pre-mRNA to mRNA splicing rates are in the range of 1 minute to 6 hours. |
| $\gamma$ | Degradation rate of mRNA    | $hr^{-1}$                     | $10^{-2} - 10^0$ | Many mRNA half-lives are in the range of a half-hour to a day.                |

@@

Given state variables representing concentrations of pre-mRNA, $u$, 
and mRNA, $s$, we have the following ordinary differential equations 
taken from [@Zeisel2011-nz] (please refer again to 
@tbl-transcription-splicing-degradation for the meaning of the variables 
and parameters),

\begin{align}
\frac{du}{dt} & = \alpha - \beta u  \label{eq-dudt}, \\
\frac{ds}{dt} & = \beta u - \gamma s  \label{eq-dsdt},
\end{align}

which proposes a mean-field model for the dynamics of transcription and 
splicing on a continuous state space. It is important to note this is only
one among an eventual ensemble of models that, with appropriate analysis
and inference, may be organized into a hierarchy of such models according
to their domains of validity relative to the scales involved in and resolution 
of observations of the relevant phenomenon [@Wells2012-qf].
Of course, in the context of single gene transcription without explicit modeling
of interactions, this ensemble of models would include the various adaptations and
extensions of the so-called random telegraph model and the experimental 
justification for its consideration 
[@Ham2020-ij;@Cao2020-xv;@Bohrer2021-qk].
As an example of a minimal extension, \eqref{eq-dudt} and \eqref{eq-dsdt} are 
usually presented with the concept that the parameter $\alpha$ could better 
account for the external inputs to the regulation of transcription of the gene 
if it were allowed to be a function of time, $\alpha(t)$. However, because this
complicates or eliminates analytical tractability, and thus maximum likelihood
inference, approximation of this function may be piecewise constant 
[@Bergen2020-pj]. We will return to 
this point later, but, for the purpose of simply illustrating the relationsips 
among analyzing, simulating, and performing inference upon the parameters of a 
dynamical system, we will assume that $\alpha$ and other parameters are 
constant values.

@@

It is common to perform a dimensional analysis of dynamical models to ensure the
units of the variables and parameters are consistent and reduce the total number
of parameters by the number of dimensions of the associated dilation group 
symmetry [@Pegoraro1973-tt]. Intuitively, this maps sets of parameters
to associated equivalence classes of similar dynamics. For reference, in addition to 
[dimensional analysis](https://en.wikipedia.org/wiki/Dimensional_analysis) and 
[nondimensionalization](https://en.wikipedia.org/wiki/Nondimensionalization), 
the procedure we make use of depends on what is frequently 
referred to as the 
[Buckingham $\Pi$ theorem](https://en.wikipedia.org/wiki/Buckingham_%CF%80_theorem) 
[@Cantwell2002-dv].

@@

| $\beta$ (hr$^{-1}$) | 1 unit of $t^*$ (hr) | 10 units of $t^*$ (hr) | 100 units of $t^*$ (hr) | 1 unit of $t^{\ast}$ | 10 units of $t^{\ast}$ | 100 units of $t^{\ast}$ |
| ------------------- | -------------------- | ---------------------- | ----------------------- | -------------------- | ---------------------- | ----------------------- |
| $10^{-1}$           | 10                   | 100                    | 1000                    | 1                    | 10                     | 100                     |
| $10^0$              | 1                    | 10                     | 100                     | 1                    | 10                     | 100                     |
| $10^1$              | 0.1                  | 1                      | 10                      | 1                    | 10                     | 100                     |
| $10^2$              | 0.01                 | 0.1                    | 1                       | 1                    | 10                     | 100                     |

@@

and define dimensionless variables and parameters as

@@

```{python}
#| label: symbolic-tsd-dudt-dimensionless-equation
#| code-fold: true
du_dt_dimless = Eq(
    beta * diff(u_star * alpha / beta, t_star),
    du_dt.rhs.subs(
        {
            u: u_star * alpha / beta,
            s: s_star * alpha / beta,
            t: t_star / beta,
            gamma: gamma_star * beta,
        }
    ),
).simplify()
du_dt_dimless
```

@@

```{python}
#| label: symbolic-tsd-dsdt-dimensionless-equation
# | code-fold: true
ds_dt_dimless = Eq(
    beta * diff(s_star * alpha / beta, t_star),
    ds_dt.rhs.subs(
        {
            u: u_star * alpha / beta,
            s: s_star * alpha / beta,
            t: t_star / beta,
            gamma: gamma_star * beta,
        }
    ),
).simplify()

@@

\begin{align}
\frac{du^{\ast}}{dt^{\ast}} & = 1 - u^{\ast}  \label{eq-dustardtstar}, \\
\frac{ds^{\ast}}{dt^{\ast}} & = u^{\ast} - \gamma^{\ast} s^{\ast},  \label{eq-dsstardtstar}
\end{align}

@@

We'll solve the dimensionless system analytically, focusing on what can be 
understood about the system's dynamics.
Given specific initial conditions $u^{\ast}(0) = u^{\ast}_0$ and $s^{\ast}(0) = s^{\ast}_0$, which
we restore to arbitrary $t^{\ast}_0$ at the end of the derivation, the 
first equation

@@

$C_1$ is of course the constant determined by the initial conditions, which can 
be shown to be $C_1 = u^{\ast}_0 - 1,$ giving us

@@

This equation describes how the concentration of pre-mRNA, in units of the 
characteristic concentration scale, $U_0 = \alpha/\beta$, evolves over the 
timescale characterized by the inverse of the splicing rate, $1/\beta$. The term $e^{-t^{\ast}}$ 
indicates that any deviation of $u^{\ast}$ from its steady state $1$ decays 
exponentially over time.

@@

The complete solutions for the dimensionless system for $\gamma^{\ast} \neq 1$ are then given by

@@

For the case $\gamma^{\ast} = 1$, we can solve \eqref{eq-dustardtstar} and 
\eqref{eq-dsstardtstar} to find the solution for $u^{\ast}$ is unchanged
and $s^{\ast}$ is then given by

$$
s^{\ast}(t^{\ast}) = 1 + (s^{\ast}_0 - 1) e^{-t^{\ast}} + (u^{\ast}_0 - 1) t^{\ast} e^{-t^{\ast}}.
$$

@@

The dimensionless **pre-mRNA dynamics** $u^{\ast}(t^{\ast})$
given in \eqref{eq-ustar-dimless} 
settles to its steady state value of 1 at rate 1, regardless of the initial 
concentration, reflecting the balance between transcription and splicing 
rates. Note again that this does not refer to an arbitrary value of $1$ but 
rather represents the equivalence class of ratios $\alpha/\beta$.

The dimensionless **mRNA dynamics** $s^{\ast}(t^{\ast})$ 
given in \eqref{eq-sstar-dimless} reveals a 
slightly more complicated balance of forces. The first transient term shows 
an exponential decay influenced by $\gamma^{\ast}$,the degradation rate 
relative to the splicing rate, when $\gamma^{\ast}$ is not precisely $1$. 
The second transient term reflects how the initial deviation in pre-mRNA 
concentration from its steady state affects mRNA levels over time. The steady 
state of $s^{\ast}$ simply reflects the balance of the splicing rate, which is 
the only source of mRNA production in this model, and the degradation rate 
also given, just as in the case of the pre-mRNA, in units of the transcription 
rate relative to the splicing rate.

@@

```{python}
# | label: transcription-splicing-degradation-simulation
import diffrax
from pyrovelocity.models import solve_transcription_splicing_model
from pyrovelocity.models import solve_transcription_splicing_model_analytical
from pyrovelocity.logging import configure_logging
from jax import numpy as jnp

@@

ts0 = jnp.linspace(0.1, 4.0, 40)
ts1 = jnp.linspace(4.0 + (10.0 - 4.0) / 20, 10.0, 20)
ts = jnp.concatenate([ts0, ts1])
initial_state = jnp.array([0.1, 0.1])
params = (1.00,)
colormap_name = "RdBu"

solution = solve_transcription_splicing_model(
    ts,
    initial_state,
    params,
)
analytical_solution = solve_transcription_splicing_model_analytical(
    ts,
    initial_state,
    params,
)
analytical_simulation_error = diffrax.Solution(
    t0=ts[0],
    t1=ts[-1],
    ts=ts,
    ys=analytical_solution.ys - solution.ys,
    interpolation=None,
    stats={},
    result=diffrax.RESULTS.successful,
    solver_state=None,
    controller_state=None,
    made_jump=None,
)

@@

plot_deterministic_simulation_trajectories(
    solution=solution,
    title_prefix="TSD Model Simulated",
    colormap_name=colormap_name,
)
```

We see that if we plot the analytical solution, we get the same result as shown in @fig-transcription-splicing-degradation-plot-analytical-trajectory-1 to within the expected error of the numerical simulation ($O(10^{-5} - 10^{-8})$).

```{python}
# | label: fig-transcription-splicing-degradation-plot-analytical-trajectory-1
# | code-fold: true
# | fig-cap: Absolute error in analytical vs simulated trajectories of pre-mRNA and mRNA concentrations over time for the transcription-splicing-degradation model.
from pyrovelocity.plots import plot_deterministic_simulation_trajectories

plot_deterministic_simulation_trajectories(
    solution=analytical_simulation_error,
    title_prefix="TSD Model Analytical-Simulation Error",
    colormap_name=colormap_name,
)
```

The same simulated trajectories from @fig-transcription-splicing-degradation-plot-simulation-trajectory-1 are shown in the phase space given by $u^{\ast} \otimes s^{\ast}$ in @fig-transcription-splicing-degradation-plot-simulation-phase-portrait-1.

@@

plot_deterministic_simulation_phase_portrait(
    solution=solution,
    title_prefix="TSD Model",
    colormap_name=colormap_name,
)
```
